Discrete Ordinates

### 2017-06-06 docprep/chapbib.tex
Updates to thesis

One of the most well known MOC codes id DeCART. DeCART is actually a coupled solver in that it utilizes different solution modalities for the radial plane and the axial direction. It uses a more complex modality in the radial plane such as MOC and then uses diffusion to transport particles axially~\citep{ref:hursinm}.

@@

Fundamentally, all reconstructions algorithms are based on a thin parallel beam of x-rays rotating axially about a patient to produce a sinogram. The inverse radon transform backprojects each row of the sinogram back to physical space. Filtering each row in Fourier space and then compositing all layers together recreates the image [CTIE]. This methodology has been extended to fan beams [CITE] and cone beams [CITE]. After the filtered backprojection, iterative reconstruction techniques can be applied to greatly improve the reconstruction quality [CITE].

@@

\section{Discrete Ordinates}\label{sec:discordlit}
The dsicrete ordinates method dates back to CITE but remains one of the most prominent solution modalities for ration transport in use today. A comprehensive review of discrete ordinate methods is given by \citet{ref:lewise}. Additional references include [CITES].

>>>>

One of the most well known MOC codes is DeCART. DeCART is actually a coupled solver in that it utilizes different solution modalities for the radial plane and the axial direction. It uses a more complex modality in the radial plane such as MOC and then uses diffusion to transport particles axially~\citep{ref:hursinm}.

@@

Fundamentally, all reconstructions algorithms are based on a thin parallel beam of x-rays rotating axially about a patient to produce a sinogram. The inverse radon transform backprojects each row of the sinogram back to physical space. Filtering each row in Fourier space and then compositing all layers together recreates the image [CTIE]. This methodology has been extended to fan beams [CITE] and cone beams [CITE]. Besides the filtered backprojection (FBP) method, iterative reconstruction techniques can also be applied to greatly improve the image reconstruction quality [CITE].

@@

\section{Discrete Ordinates}\label{sec:discordlit}
The dsicrete ordinates method dates back to CITE but remains one of the most prominent solution modalities for radiation transport in use today. A comprehensive review of discrete ordinate methods is given by \citet{ref:lewise}. Additional references include [CITES].

### 2017-06-06 docprep/chapimpl.tex
Updates to thesis

The code is available from the GitHub repository. At the time of this writing, the repository is located at \texttt{https://github.com/eNorris/thesis}.

>>>>

The code is available from a GitHub Git repository. At the time of this writing, the repository is located at \texttt{https://github.com/eNorris/thesis} and is only available subject to special request and license agreement. An executable version is available.

### 2017-06-06 docprep/chapintro.tex
Updates to thesis

Though concerns have been raised in the medical community about the risks associated with CT, the techniques used to quantify dose are very simple and not patient specific. The dose-length-product (DLP) is used as a measure to quantify the effective dose a patient receives. However, the DLP must be modified by a factor accounting for may variations amongst patients such as age, gender, and size as well as the kVp used in the procedure. Currently, dose estimation relies on \textit{a priori} computation verified with a standardized benchmark.

>>>>

Though concerns have been raised in the medical community about the risks associated with CT, the techniques used to quantify dose are very simple and not patient specific. The dose-length-product (DLP) is used as a measure to quantify the effective dose a patient receives. However, the DLP must be modified by a factor accounting for many variations amongst patients such as age, gender, and size as well as the kVp used in the procedure. Currently, dose estimation relies on \textit{a priori} computation verified with a standardized benchmark.

### 2017-06-06 docprep/chapmeth.tex
Updates to thesis

The user must provide cross section data in the form of an AMPX formatted binary file (Format detaisl are provided in Section~\ref{sec:xsparse}). The cross section data file contains group structure information. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight the energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries us a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

@@

The group structure cannot be arbitrarily selected by DOCOTRS, instead it is determined by the external cross section data file loaded by the user. The energy group structure should be selected appropriately for the problem. The cross sections currently included with DOCTORS are those available from the SCALE 6.2 distribution. Those cross sections are designed for light water reactor analysis and are not well suited to medical physics applications. This is currently one of the greatest limitations of DOCOTRS. The SCALE cross section data also contains neutron data which is useless for the CT dosimetry DOCTORS performs. More details of the cross section parsing and generation are included in Section~\ref{sec:xsparse} and~\ref{sec:xsgen} respectively.

@@

The problem domain is split into an evenly spaced Cartesian grid. Uniform grid spacing is not required by discrete ordinate methods though it is helpful [CITE], but is necessarily the case with CT voxel phantoms.

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient:
\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}.
\end{equation}
To a first order approximation, a partial derivative is computed
\begin{equation} \label{eq:deriv_1}
\frac{\partial f(\zeta)}{\partial \zeta} \approx \frac{f(\zeta+\Delta \zeta/2) - f(\zeta + \Delta \zeta/2)}{\Delta \zeta}
\end{equation}
with respect to $\zeta$. Combining Eq.~\ref{eq:grad} and~\ref{eq:deriv_1}, the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as:
\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi.
\end{split}
\end{equation}
The flux variables $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

@@

Treatment for the $\theta$ variable is similar to that of $\varphi$ except the polar angle is required instead of the azimuthal angle. The polar angle of an arbitrary vector $\boldsymbol{\xi}$ can be computed using \begin{equation}\label{eq:thetacos}
\cos(\theta_\xi) = \frac{\xi_z}{\sqrt{\xi_z^2 + \xi_y^2 + \xi_z^2}}.
\end{equation}
This is used to compute the polar angle for both $\boldsymbol{s}$ and $\boldsymbol{c}$. The difference between the polar angles of the two vectors 
\begin{equation}\label{eq:thetacos2}
\zeta_\theta = \theta_s - \theta_c
\end{equation}
is used to evaluate
\begin{equation}\label{eq:thetacoscon}
|\zeta_\theta| < \frac{\theta}{2}.
\end{equation}
If the condition is not met, the ray is rejected and no raytrace is performed. Figure~\ref{fig:fan_rejection} shows a simple example demonstrating cases when a voxel would be accepted or rejected.

@@

The alternative way to compute the dose is directly from the local energy deposition. The total deposition, $\mathcal{E_{T,i}}$ for voxel $i$ is
\begin{equation}\label{eq:dept}
\mathcal{E}_{T,i} = \mathcal{E}_{a,i} + \mathcal{E}_{s,i}
\end{equation} 
where $\mathcal{E_{a,i}}$ is the energy deposition from absorption and $\mathcal{E_{s,i}}$ is the energy deposition from scatter. The absorption deposition is
\begin{equation}\label{eq:depa}
\mathcal{E}_{a,i} = \sum_{g=0}^{G-1} \Sigma_{a,i}^g \varphi_i^g V_i \bar{E}^g F_1
\end{equation}
where $\bar{E}$ is the average energy (in MeV) of the $g^{th}$ group and $F_1$ is a unit conversion factor that converts MeV to Joules. Equation~\ref{eq:depa} assumes total local energy deposition in the voxel of interest. The energy deposition from scatter is
\begin{equation}\label{eq:deps}
\mathcal{E}_{s,i} = \sum_{g=0}^{G-1} \sum_{h=g-1}^{G-1} \left[ \Sigma_{s,i}^{g,h} \varphi_i^g V_i (\bar{E}^g-\bar{E}^h) F_1 \right]
\end{equation}
making the total deposition
\begin{equation}\label{eq:dept2}
\mathcal{E}_{T,i} = V_i F_1 \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g-1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right]
\end{equation}
The mass of the voxel, $M_i$ is then computed
\begin{equation}
M_i = \rho_i V_i F_2
\end{equation}
where $\rho_i$ is the density in g/cm$^3$ and $F_2$ is a conversion factor to convert grams to kilograms. The dose (in Gy) to the voxel is then
\begin{equation}\label{eq:dosedep}
D_i = \frac{\mathcal{E}_{T,i}}{M_i} = \frac{F_1}{\rho_i F_2} \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g-1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right].
\end{equation}

>>>>

The user must provide cross section data in the form of an AMPX formatted binary file (Format detaisl are provided in Section~\ref{sec:xsparse}). The cross section data file contains group structure information. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight the energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries use a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

@@

The group structure cannot be arbitrarily selected by DOCOTRS, instead it is determined by the external cross section data file loaded by the user. The energy group structure should be selected appropriately for the problem. The cross sections currently included with DOCTORS are those available from the SCALE 6.2 distribution. Those cross sections are designed for light water reactor analysis and are not well suited to medical physics applications. This is one of the limitations of the current version of DOCOTRS. The SCALE cross section data also contains neutron data which is useless for the CT dosimetry DOCTORS performs. More details of the cross section parsing and generation are included in Section~\ref{sec:xsparse} and~\ref{sec:xsgen} respectively.

@@

The problem domain is split into an evenly spaced Cartesian grid. Uniform grid spacing is not required by discrete ordinate methods though it is helpful and can speed up the computation[CITE].

@@

To solve the fully discretized form of the Linear Boltzmann Equation given in Eq.~\ref{eq:boltz_i}, the gradient operator must first be computed numerically. Recall that $\Omega_a$ can be written in vector notation using its cosine components as defined in Eq.~\ref{eq:omega_cos}. Also recall the definition of the gradient:
\begin{equation} \label{eq:grad}
\nabla f(x, y, z) = \frac{\partial f(x, y, z)}{\partial x} \hat{i} + \frac{\partial f(x, y, z)}{\partial y} \hat{j} + \frac{\partial f(x, y, z)}{\partial z} \hat{k}.
\end{equation}
To a first order approximation, a partial derivative is computed
\begin{equation} \label{eq:deriv_1}
\frac{\partial f(\zeta)}{\partial \zeta} \approx \frac{f(\zeta+\Delta \zeta/2) - f(\zeta - \Delta \zeta/2)}{\Delta \zeta}
\end{equation}
with respect to $\zeta$. Combining Eq.~\ref{eq:grad} and~\ref{eq:deriv_1}, the $\hat{\Omega} \cdot \nabla \psi$ term can be rewritten as:
\begin{equation} \label{eq:spatial_1}
\begin{split}
\Omega \cdot \nabla \psi 
& = <\mu, \eta, \xi> \cdot <\frac{\partial \psi}{\partial x}, \frac{\partial \psi}{\partial y}, \frac{\partial \psi}{\partial z}> \\
& =
\frac{\partial \psi}{\partial x}\mu + \frac{\partial \psi}{\partial y}\eta + \frac{\partial \psi}{\partial z}\xi \\
& \approx 
\frac{\psi(x + \Delta x/2, y, z) - \psi(x - \Delta x/2, y, z)}{\Delta x} \mu \\
&+ 
\frac{\psi(x, y + \Delta y/2, z) - \psi(x, y - \Delta y/2, z)}{\Delta y} \eta \\
&+ 
\frac{\psi(x, y, z + \Delta z/2) - \psi(x, y, z - \Delta z/2)}{\Delta z} \xi.
\end{split}
\end{equation}
The flux variables $\psi_{i,a,x,in}^g$, $\psi_{i,a,x,out}^g$, $\psi_{i,a,y,in}^g$, $\psi_{i,a,y,out}^g$, $\psi_{i,a,z,in}^g$, and $\psi_{i,a,z,out}^g$ are \textit{surface} averaged flux values while $\psi_{i,a}^{g}$ is a \textit{volume} averaged flux value.

@@

Treatment for the $\theta$ variable is similar to that of $\varphi$ except the polar angle is required instead of the azimuthal angle. The polar angle of an arbitrary vector $\boldsymbol{\xi}$ can be computed using \begin{equation}\label{eq:thetacos}
\cos(\theta_\xi) = \frac{\xi_z}{\sqrt{\xi_x^2 + \xi_y^2 + \xi_z^2}}.
\end{equation}
This is used to compute the polar angle for both $\boldsymbol{s}$ and $\boldsymbol{c}$. The difference between the polar angles of the two vectors 
\begin{equation}\label{eq:thetacos2}
\zeta_\theta = \theta_s - \theta_c
\end{equation}
is used to evaluate
\begin{equation}\label{eq:thetacoscon}
|\zeta_\theta| < \frac{\theta}{2}.
\end{equation}
If the condition is not met, the ray is rejected and no raytrace is performed. Figure~\ref{fig:fan_rejection} shows a simple example demonstrating cases when a voxel would be accepted or rejected.

@@

The alternative way to compute the dose is directly from the local energy deposition. The total deposition, $\mathcal{E_{T,i}}$ for voxel $i$ is
\begin{equation}\label{eq:dept}
\mathcal{E}_{T,i} = \mathcal{E}_{a,i} + \mathcal{E}_{s,i}
\end{equation} 
where $\mathcal{E_{a,i}}$ is the energy deposition from absorption and $\mathcal{E_{s,i}}$ is the energy deposition from scatter. The absorption deposition is
\begin{equation}\label{eq:depa}
\mathcal{E}_{a,i} = \sum_{g=0}^{G-1} \Sigma_{a,i}^g \varphi_i^g V_i \bar{E}^g F_1
\end{equation}
where $\bar{E}$ is the average energy (in MeV) of the $g^{th}$ group and $F_1$ is a unit conversion factor that converts MeV to Joules. The average group energy is assumed to be the average between the two bounding energies of the group. Equation~\ref{eq:depa} assumes total local energy deposition in the voxel of interest. The energy deposition from scatter is
\begin{equation}\label{eq:deps}
\mathcal{E}_{s,i} = \sum_{g=0}^{G-1} \sum_{h=g-1}^{G-1} \left[ \Sigma_{s,i}^{g,h} \varphi_i^g V_i (\bar{E}^g-\bar{E}^h) F_1 \right]
\end{equation}
making the total deposition
\begin{equation}\label{eq:dept2}
\mathcal{E}_{T,i} = V_i F_1 \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right]
\end{equation}
The mass of the voxel, $M_i$ is then computed
\begin{equation}
M_i = \rho_i V_i F_2
\end{equation}
where $\rho_i$ is the density in g/cm$^3$ and $F_2$ is a conversion factor to convert grams to kilograms. The dose (in Gy) to the voxel is then
\begin{equation}\label{eq:dosedep}
D_i = \frac{\mathcal{E}_{T,i}}{M_i} = \frac{F_1}{\rho_i F_2} \sum_{g=0}^{G-1} \varphi_i^g \left[\Sigma_{a,i}^g \bar{E}^g + \sum_{h=g+1}^{G-1} \left\{ \Sigma_{s,i}^{g,h}  (\bar{E}^g-\bar{E}^h) \right\} \right].
\end{equation}

### 2017-06-06 docprep/norris_phd_dissertation.tex
Updates to thesis

Computed tomography (CT) has become pervasive in medical diagnostics as improved imaging techniques and processing algorithms provide higher quality information to doctors. However, the exponentially increasing usage of CT has raised concerns regarding long term radiological risks.

>>>>

Computed tomography (CT) has become pervasive in medical diagnostics as improved imaging techniques and processing algorithms provide higher quality information to doctors. However, the exponentially increasing usage of CT has raised concerns regarding long term low-dose radiological risks. Currently, the dose to patients is computed using Monte Carlo methods and experimental tests. In other areas of radiation transport, deterministic codes have been shown to be much faster than Monte Carlo codes.

Currently, no deterministic methodology exists to automatically generate a spatially distributed dose profile from a CT voxel phantom. This work proposes a new code, Discrete Ordinate CT Organ Dose Simulator (DOCTORS) which utilizes a GPU accelerated raytracer and discrete ordinate solver to compute photon flux in the patient. The flux is then converted to dose.

The DOCTORS code was benchmarked against MCNP6 and found to have very good agreement using both a water phantom and a realistic patient phantom. DOCTORS was also found to be much faster than MCNP6, even more so with GPU acceleration.

GPU
